---
title: "International breakpoint analysis"
output: html_notebook
---

The aim of this is to *quickly* test the following claim made by Gerry in a recent presentation:

`Mortality trend change is real and substantial. Scotland, England & Wales, USA worst affected. Other countries not affected at all. `

My sense from looking at international data is that the first part of the claim might be true, but the latter part is not, and instead the stalling has been international.

However I've not yet done a segmented regression/breakpoint analysis of international life expectancy data to (dis)confirm this. This document aims to quickly perform this analysis and work out if there has been a similar breakpoint in many/most HMD countries.


```{r}
rm(list = ls())

pacman::p_load(
  tidyverse, 
  plotly,
  HMDHFDplus,
  segmented
)

```


```{r}
baseloc <- "N:/Jon/hmd_data/hmd_countries/"

codes <- dir(baseloc)

hmd_e0 <- data_frame(
  code = codes
) %>% 
  mutate(
    e0 = map(
      code,
      function(x) {
        read_table(
          file.path(
            baseloc, x, "STATS/e0per.txt"
          ),
          skip = 2,
          col_types = "iddd"
          
        )
      }
    )
  ) %>% 
  unnest() %>% 
  gather(key = "gender", value = "e0", Female:Total)
```


Let's look at 1990 onwards.

```{r}

modblock <- hmd_e0 %>% 
  filter(Year >= 1990) %>% 
  filter(gender != "Total") %>% 
  group_by(code, gender) %>% 
  nest() 

n_models <- dim(modblock)[1]

modblock <- modblock %>% 
  mutate(mod_results = vector("list", length = n_models))

for (i in 1:n_models){
  this_data <- modblock$data[[i]]
  this_lm <- lm(e0 ~ Year, data = this_data)
  
  this_seg <- try(
    segmented(this_lm, seg_rate.z = ~Year, psi = NA, control = seg.control(K=1, stop.if.error = F)),
    TRUE
  )
  
  modblock$mod_results[[i]] <- this_seg
}


```

Now, for those datasets where a breakpoint can be identified, let's pull these out 


```{r}

get_psi <- function(identified, mod_results){
  if(!identified){
    return(NA_real_)
  } 
  return(mod_results$psi[,"Est."])
}

get_se <- function(identified, mod_results){
  if(!identified){
    return(NA_real_)
  } 
  return(mod_results$psi[,"St.Err"])
}


modblock_results <- modblock %>% 
  mutate(is_identified = map_lgl(mod_results, ~class(.x)[1] != 'try-error')) %>% 
  mutate(seg_break = map2_dbl(
    is_identified, mod_results,
    get_psi
    )
  ) %>% 
  mutate(seg_se = map2_dbl(
    is_identified, mod_results,
    get_se
  )) %>% 
  select(code, gender, seg_break, seg_se)

```

And now to visualise

```{r}
modblock_results %>% 
  ggplot(aes(y = code)) + 
  geom_point(aes(x = seg_break)) +
  geom_errorbarh(aes(xmin = seg_break - 2 * seg_se, xmax = seg_break + 2 * seg_se)) + 
  facet_wrap(~gender)

```

 Now what if I attempt a 2 break point version? 
 
 
```{r}

modblock2 <- hmd_e0 %>% 
  filter(Year >= 1985) %>% 
  filter(gender != "Total") %>% 
  group_by(code, gender) %>% 
  nest() 

n_models <- dim(modblock2)[1]

modblock2 <- modblock2 %>% 
  mutate(mod_results = vector("list", length = n_models))

for (i in 1:n_models){
  this_data <- modblock2$data[[i]]
  this_lm <- lm(e0 ~ Year, data = this_data)
  
  this_seg <- try(
    segmented(this_lm, seg_rate.z = ~Year, psi = NA, control = seg.control(K=2, stop.if.error = F)), # note K=2
    TRUE
  )
  
  modblock2$mod_results[[i]] <- this_seg
}


```
 
```{r}

get_psi <- function(identified, mod_results){
  if(!identified){
    return(NA_real_)
  } 
  return(mod_results$psi[1,"Est."])
}

get_se <- function(identified, mod_results){
  if(!identified){
    return(NA_real_)
  } 
  return(mod_results$psi[1,"St.Err"])
}

get_psi2 <- function(identified, mod_results){
  if(!identified){
    return(NA_real_)
  } 
  return(mod_results$psi[2,"Est."])
}

get_se2 <- function(identified, mod_results){
  if(!identified){
    return(NA_real_)
  } 
  return(mod_results$psi[2,"St.Err"])
}


modblock2_results <- modblock2 %>% 
  mutate(is_identified = map_lgl(mod_results, ~class(.x)[1] != 'try-error')) %>% 
  mutate(
    seg_break = map2_dbl(is_identified, mod_results, get_psi),
    seg_break2 = map2_dbl(is_identified, mod_results, get_psi2)
  ) %>% 
  mutate(
    seg_se = map2_dbl(is_identified, mod_results,get_se),
    seg_se2 = map2_dbl(is_identified, mod_results,get_se2),
         ) %>% 
  select(code, gender, seg_break, seg_se, seg_break2, seg_se2)

```

And now to visualise this 


```{r}
#order by seg_break2 for females 

reordered_codes <- modblock2_results %>% 
  filter(gender == "Female") %>% 
  mutate(code = fct_reorder(code, seg_break2)) %>% 
  pull(code)

modblock2_results %>% 
#  mutate(code = factor(code, levels = levels(reordered_codes))) %>% 
  ggplot(aes(y = code)) + 
  geom_point(aes(x = seg_break), colour = "blue") +
  geom_errorbarh(aes(xmin = seg_break - 2 * seg_se, xmax = seg_break + 2 * seg_se), colour = "blue") + 
    geom_point(aes(x = seg_break2), colour = "red") +
  geom_errorbarh(aes(xmin = seg_break2 - 2 * seg_se2, xmax = seg_break2 + 2 * seg_se2), colour = "red") + 
  facet_wrap(~gender)

```